---
title: "Merging"
output: html_document
---

In addition to the data manipulation gymnastics we learned last week, there are two other core proficiencies you should have in your data wrangling toolkit: *merging* two datasets together, and *reshaping* datasets into a different format. Let's start with merging.

As always, I start out by loading some relevant libraries:
```{r}
library(data.table)
```
<br> 

## Merging
Let's say I have two datasets about students' test scores, named `scores_exam_1` and `scores_exam_2`:
```{r}
## note: this is how you make a data.table from scratch! The values to the left of the "=" are column names, the vectors to the right are the values. You should recognize most of these functions, except "sample". Type ?sample into your console to learn what it does.
scores_exam_1 <- data.table(class_section=c(rep("A", 5), rep("B", 6)),
                            student_id=c(1:5, 1:6),
                            score_1=sample(70:100, 11, replace=T)) # randomly generate some plausible test scores
print(scores_exam_1)

scores_exam_2 <- data.table(class_section=c(rep("A", 5), rep("B", 6)),
                            student_id=c(1:5, 1:6),
                            score_2=sample(70:100, 11, replace=T))
print(scores_exam_2)
```

<br>
I'd like to make these into a single dataset so that I can (for example) take the difference between the two test scores for each student, or plot one test score against the other. How can I do this? Using the `merge()` command, of course!

Let's take a look back up at our datasets. Each one has three columns: `student_id`, `class_section`, and some sort of "score" variable. The `student_id` and `class_section` variables are the same in each dataset, while the "score" variable has both a *different name* and *different values*. What we'd like to end up with is a dataset with *four* variables: `student_id`, `class_section`, `score_1`, and `score_2`:
```{r, echo=F}
merge(scores_exam_1, scores_exam_2, by=c("class_section", "student_id"))
```

`merge()` takes three main arguments: `x`, `y`, and `by`[^1]. `x` and `y` are the datasets you want to merge together, while `by` is a vector of the column names the two datasets have in *common*. This is how R knows which row to line up with which in any given dataset. So, to merge my two datasets together, I would do: 

[^1]: Note: No relation to the`by` argument you use within `data.table` commands!
```{r}
merge(scores_exam_1, scores_exam_2, by=c("class_section", "student_id"))
```
<br>
Notice that I need to include *both* `class_section` and `student_id` in order to effectively merge my dataset, because it's only the *combination of those two* that uniquely identify a student. Let's see what happens when I try to merge with just one of those columns:
```{r, error=T}
merge(scores_exam_1, scores_exam_2, by=c("student_id"))
```
<br>
Whoa, that's not what we wanted! That's some weird mix of stacking and merging and it's not useful. Be careful! 

There's one more argument type you should be aware of in `merge()`: the `all`, `all.x`, and `all.y` arguments. To understand why they're useful, let's say that a new class section was added between exams 1 and 2, so you `scores_exam_2` has a class section, "C", that `scores_exam_1` lacks:
```{r}
scores_exam_2 <- data.table(class_section=c(rep("A", 5), rep("B", 6), rep("C", 4)),
                            student_id=c(1:5, 1:6, 1:4),
                            score_2=sample(70:100, 15, replace=T))
print(scores_exam_2)
```
<br>
By default, the `merge` command will keep *only rows that are found in both datasets*. So running our original `merge` command will get us the same outcome we had before:
```{r}
merge(scores_exam_1, scores_exam_2, by=c("class_section", "student_id"))
```
<br>
But what if we wanted to keep *all* the rows from `scores_exam_2`, and just fill in with missing values for exam 1 scores? We could do that by instructing `merge` to keep *all* the rows from the *y* dataset, by setting the argument `all.y` equal to `TRUE`:
```{r}
merge(scores_exam_1, scores_exam_2, by=c("class_section", "student_id"), all.y=TRUE)
```
<br>
Now let's imagine that additionally, there's a student in section A (id number 6) who dropped after the first exam, so he doesn't show up in dataset 2:
```{r}
scores_exam_1 <- data.table(class_section=c(rep("A", 6), rep("B", 6)),
                            student_id=c(1:6, 1:6),
                            score_1=sample(70:100, 12, replace=T))

print(scores_exam_1)
```
<br>
You can include *just* this extra student (and nobody in section C from the other dataset), by running the `merge` command with `all.x` equal to `TRUE`. Or, if you wanted to keep all the extra rows in *both* datasets, you could run with both `all.x` and `all.y` equal to `TRUE`. It's common enough to want to do this that `merge` has an `all` argument that, when set to `TRUE`, is equivalent to setting both `all.x` and `all.y` equal to  `TRUE`. Try it yourself!


