---
title: "Objects, Data Types, ID Arrays"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is our first real coding lecture. In this and all future such lectures, I highly highly recommend that you work through examples with me as we go along.

## Basic Data Types
Remember from our last lecture that you can use R as a calculator: 
```{r}
# addition
4+5  
```
```{r}
# subtraction
19-17 
```
```{r}
# multiplication
5*3 
```
```{r}
# division
9/38 
```
```{r}
# exponentiation
2^4 
```

<br>
And that you can use it to display and manipulate strings:
```{r}
# display text in the console
print("Hello, World!") 
# select the 3rd-7th characters of the phrase "Hello, World!"
substr("Hello, World!", 3, 7) 
```

<br> 
And that you can test the equality, inequality, or relative size of two statements or numbers:
```{r}
# test the equality of numbers
4 == 2+2 
# test the INequality of numbers
4 != 2+2 

# test the equivalence of phrases
"Hello"=="hello" 
# test the INequivalence of phrases
"Hello"!="hello" 

# size comparisons
3>2 
3<2
3<3

```

<br> 
These three types of operation mirror the three basic ways that data is stored in R. **Numeric** data types are numbers like `7` or `-2` or `3.14159`. **Character** data types, also called **strings**, are letters or phrases enclosed by quotes, like `"Hello, World!"` or `"This is a string"`, or `"3.14159"`. (Yes, if you put a number in quotes it becomes a string!) **Logical** data types, also known as **booleans**, are `TRUE` or `FALSE` values, usually generated as a result of *testing the equivalence* of two objects. 

Get to know these data types well-- programming is all about manipulating them in intricate ways.


## Objects and Variables
So far, you're probably less-than-impressed that R knows how to add two numbers together. Fair. Programming languages like R really start to get powerful when you realize that you can **store** values in named objects called **variables**. 

Let's say you're at home, and you want to bake some brownies while you work on your Intro to Data Science homework. Unfortunately, your brownie pan is small, so you want to make a batch 2/3 the size it says in the recipe. You could type in the amount of flour you'd need to measure:

```{r}
# the recipe calls for 2 cups of flour, so the number of cups I need is:
2*(2/3)
```

And then type in the amount of butter:
```{r}
# the recipe calls for 8 tablespoons of butter, so the number of tablespoons I need is:
8*(2/3)
```

And then the amount of sugar:
```{r}
# the recipe calls for 1.5 cups of sugar, so the number of cups I need is:
1.5*(2/3)
```

...but you see already how this is getting tedious. We want to do the same thing over and over (multiply a number by 2/3), but changing (or, some might say, *varying*) the value of the number we're multiplying. This is exactly what a **variable** is designed for. 

If you're following along with this code in RStudio, look at `Environment` section in the top right corner of the screen. At the moment it should say "Environment is empty"-- what that means is that we haven't *stored* anything as a variable yet. Now, run this:

```{r}
recipe_amount <- 2
```
You should notice two things about running this code. First, it didn't print any *output* to the console. Second, there should now be an entry in the "Environment" section, with the name `recipe_amount` and a value of 2:

[insert screenshot]

So, what have we done? We haven't multiplied anything by anything, but we've created a **variable** named `recipe_amount`, set its value equal to 2 (the number of cups of flour specified in the recipe), and *saved* it in the R environment. The `<-` operator is what we use in R to **assign** a value to a variable name. (In case it's unclear: that's not a single symbol, it's a left caret followed by the minus sign.)

Now that we have this variable, we can do math (or anything else) to it just the same as we would to the initial number. Here, we want to multiply it by 2/3, so we can write:

```{r}
recipe_amount * (2/3)
```

As you can see, this gives us the correct number of cups of flour to use: 1 & 1/3 (aka 1.3333...).

At this point, we have a two-line script that looks like this:
```{r, eval=F}
recipe_amount <- 2
recipe_amount * (2/3)
```

Now, if we want to figure out how many tablespoons of butter we need, all we have to do is change the 2 to an 8 and run those two lines of code again, and the computer will print out the correct number. Try it out!

What if, in two weeks, you buy a much bigger baking pan, so now you instead want to multiply the original recipe by 2.5? You could change your two-line script to a three-line script that looks like this:

```{r}
recipe_amount <- 2
recipe_multiplier <- 2.5
recipe_amount * recipe_multiplier
```
Now you've expanded your script to *two* variables, and you can change either of them as needed to get your ingredient quantities!

This may seem like a silly example--it would still be a bit tedious to change the value of the `recipe_amount` variable over and over again, and it would probably be easier just to use an actual calculator, but there are two things I want you to take away from this:

1. It's possible to *store* values in memory as named variables and manipulate them later;
2. It would be super convenient if you could convert *all* the ingredient amounts in one go, like if you could sort of "loop" through the list of ingredients and multiply each one by your `recipe_multiplier` variable. Hmmm (*foreshadowing*)...


<br>
Also note: You can use variables to create other variables. So, if I wanted to save the final amount of some ingredient, I could do:
```{r}
final_quantity <- recipe_amount * recipe_multiplier
```
<br>
You can make string variables or boolean variables as well as numeric variables:
```{r}
# String variables
string_variable <- "Hello, World!"
print(string_variable)
substr(string_variable, 3, 7)

# Boolean variables
boolean_variable <- 2==4
boolean_variable
```
<br>

Notice that I can name variables whatever I want. I called our first variable `recipe_amount`, but I could have called it `goldfish` and the code would have understood it just fine:
```{r}
# generate a variable named "goldfish"
goldfish <- 8 
# multiply that variable by 2/3
goldfish*(2/3)
```
<br>
Also notice that R is extremely sensitive to small typos, like letter ordering or capitalization:
```{r, error=T}
print(goldfish)
print(Goldfish)
print(godlfish)
```
<br>
Naming things whatever we want is exciting, but we should use this power responsibly. Variable names (and object names more broadly), should be clear, concise, and descriptive of the thing they represent. If you name your objects well enough, it should really cut down on the number of comments you need to write. For example, in the section below on 1D arrays, you'll notice that instead of writing code like this:

```{r, eval=F}
# make a numeric vector
c(1, 3, 8)
```

I create the numeric vector and save it as an object with a reasonable name:
```{r, eval=F}
numeric_vector <- c(1, 3, 8)
```

This conveys the same information more concisely and replicably. See the coding best practices page for more details. 

I've thrown the word "object" around a few times without really describing what it is. An **object** is anything that you can manipulate in a programming language. They're usually named, and you can see them in your `Environment` panel. Variables are objects, but so are lots of other things (like arrays, which we'll learn about in the next section). 

Take a break! 

[insert gif]

## 1-Dimensional Arrays

So far we've just looked at objects that represent single things, like the number 8 or the word "cat" or the value "TRUE". What if we instead want a *collection* of things, like a list of numbers or words or true/false values (such as, perhaps, a list of recipe quantities)? Well, we'd use an array. 

**"Array"** is a generic term for a group of things. Arrays can have many different formats, dimensions, and names. We're going to be focusing on a type of 1-dimensional array in R called a **vector**. (1-dimensional means that there's just one list of things-- a 2-dimensional array would be like a spreadsheet or a dataset, we'll get to those later.) 

### Making Vectors

#### `c()`
There are a few different ways to make vectors in R; the most manual way is using a fuction called `c()`. `c()` takes all the elements you give it (separated by commas) and *combines* elements into a vector (the `c` is for "combine"). Let's look at a few examples:

```{r}
numeric_vector <- c(1, 3, 8)
print(numeric_vector)

string_vector <- c("cat", "dog", "horse")
print(string_vector)

boolean_vector <- c(F, T, T, F)  # note: you can just write "T" or "F" instead of "TRUE" and "FALSE"; it achieves the same thing.
print(boolean_vector)
```

<br>
Notice that all the *members* of each of these vectors is the same type. For vectors, all elements **must** be the same type. If you try to make a vector with multiple types, R will *coerce* them to all be the same:

```{r}
# strings "beat" numerics
c(3, "cat", 4, "dog") 
# numerics "beat" logicals (in binary arithmetic, 1 equals "True" and 0 equals "False", hence the conversions)
c(7, T, F, 2) 
# strings "beat" logicals
c("Hello", F, "World!", T, T) 
```
<br>

#### Other Ways to Make Vectors
There are a few other ways to make vectors that are useful to know. If you want a series of numbers in sequence, you can use the `:` operator and then you just have to specify the first and last number you want:
```{r}
# vector of integers from 1 to 10
1:10 
# integers from 57 to 65
57:65 
# it can go backwards, too!
10:1 
# it's fine with decimal numbers, but the interval between values will always be 1
5.5:8.5 
```
<br> 
If you want a series of numbers at intervals other than 1, use the `seq()` function. It needs three values: your starting number, your final number, and the interval between values:
```{r}
# this is the same as "1:10"
seq(1, 10, 1) 
# 1:10, but only keeping every other number
seq(1, 10, 2) 
# 10:100, only keeping every tenth number
seq(10, 100, 10) 
```
<br> 
If you want to repeat the same set of values many times, use the `rep()` function. It needs two things: the vector of values you want to repeat, and how many times you want to repeat them:
```{r}
# repeat the sequence 1:5 three times
rep(1:5, 3) 
# repeat this two-element string vector twice
rep(c("Hello", "World!"), 2) 
# repeat this three-element boolean vector four times
rep(c(F, F, T), 4) 
```

### Subsetting Vectors
<br> 
Let's say you make a vector:
```{r}
example_vector <- c("Oxford", "London", "Cambridge", "Manchester", "Exeter", "Leeds")
print(example_vector)
```
This vector is six elements long (which you can verify by running `length(example_vector)`). R records both the fact that `example_vector` contains these six values and *which position* each value is in. The position of a value in the vector is called its **index**. So "Oxford" would be in position #1 of the vector, "London" in position #2, and so on and so forth.  

You can use these index vaues to *subset* your vector. R uses square brackets `[]` for indexing operations. If you wanted to know, say, what the 5th element of the vector was, you would type:

```{r}
example_vector[5]
```

If you wanted to know what the 2nd *through* 5th elements were, you could use the range operator `:` that we learned earlier:

```{r}
example_vector[2:5]
```

And similarly you can use vector notation if you want to know the 1st, 3rd, and 5th values:

```{r}
example_vector[c(1, 3, 5)]
```


What if you have the opposite question: you know that "Cambridge" is in your vector somewhere, and you want to know what position it's in? You would use the `which()` function, which takes as input a *logical statement* like the ones we've been learning. That code would look like this:

```{r}
which(example_vector=="Cambridge")
```

**Exercises**: Think about the following vector:
```{r}
exercise_vector <- c("Cambridge", "Oxford", "Leeds", "Cambridge")
```

* Use code to determine the index value for "Leeds".
* What would happen if you tried to use `which` to determine the index value for "Cambridge"? Make a guess before you run the code. 


#### Note on Indexing and Numeric Vectors
Think about the numeric vector we made at the beginning of this lesson:
```{r}
print(numeric_vector)
```

Here, the elements of the vector are numbers, not strings, but the way *indexing* works is exactly the same. So the first element of the vector is the number 1, the second element is the number 3, and the third element is the number 8. This can be hard to keep track of at times because both the elements of the vector and the indices of the vector are numbers (and those numbers are not the same). Guess what the output will be when you run the following lines of code, then check to see if you're right:

```{r, eval=F}
numeric_vector[1]
which(numeric_vector==1)

numeric_vector[2]
which(numeric_vector==2)

numeric_vector[3]
which(numeric_vector==3)

numeric_vector[8]
which(numeric_vector==8)

```



### Lengthening, Shortening, and Combining Vectors

#### Lengthening Vectors
Remember what our `example_vector` looks like:

```{r, echo=F}
print(example_vector)
```

Shoot, I forgot a city! I also wanted to include "Birmingham" in this list. I could re-create my entire vector with the additional element:

```{r}
c("Oxford", "London", "Cambridge", "Manchester", "Exeter", "Leeds", "Birmingham")
```


...but that's quite time-consuming and annoying. It would be better if there were a way for me to easily just tack another element on to the end of the list. It turns out, that's easy to do-- and in fact, it uses the same function we used to create the vector in the first place-- the "combining" function, `c()`! 

To **append** the element "Birmingham" to our `example_vector`, we would do: 

```{r}
c(example_vector, "Birmingham")
```
<br>

#### Important Note: Execution vs Assignment 

What is the difference between these two lines of code?

```{r, eval=F}
## Example Line 1
c("cheese", "chips", "popcorn")

## Example Line 2
snack_vector <- c("cheese", "chips", "popcorn")
```

As you hopefully remember from the section above about creating variables, the key difference here is one of **assignment**-- that is, of saving values to memory. The first line tells R to create a vector, print it in the console screen, and then 'forget' about it immediately. The second line tells R to create a vector, *not* print it to the console, and save it in memory as an object named `snack_vector`. The first line is just **executing** the "create a vector" function, the second one is both executing that function and *assigning the output* to an object name. 

This same logic applies to operations that *modify existing objects*-- if you don't run the special `<-` operator to *assign* the change to an object, the change won't get saved anywhere. Let's think of our example above, in which we want to append the element "Birmingham" to the vector `example_vector`. Consider these three commands:

```{r}
## Command 1: execute, but don't assign
c(example_vector, "Birmingham")
print(example_vector)
```

Here, we've printed our new, expanded vector to the screen, but the *saved* `example_vector` is unchanged. 

```{r}
## Command 2: assign new vector to a new object name
longer_example_vector <- c(example_vector, "Birmingham")
print(example_vector)
print(longer_example_vector)
```

Here, because we assigned the command to a *new* object name, `example_vector` remains unchanged and we have a new vector, `longer_example_vector`, that contains the expanded vector.

```{r}
## Command 3: *update* the existing vector
example_vector <- c(example_vector, "Birmingham")
print(example_vector)
```

Here, because we assigned the command to the *same* object name, we've *overwritten* our original value of `example_vector`, and `example_vector` itself is longer now. This type of operation, where we change the value of an existing object, is called **updating** the object.

**Exercise: Now that we've updated `example_vector`, what will `example_vector` look like if we run this code?

```{r, eval=F}
c(example_vector, "Aberdeen")
```

What about this code?

```{r, eval=F}
new_vector <- c(example_vector, "Aberdeen")
```

And this code?
```{r, eval=F}
example_vector <- c(example_vector, "Aberdeen")
```

#### Shortening Vectors

We've successfully updated our `example_vector` to include "Birmingham":
```{r, echo=F}
print(example_vector)
```

What if we now want to *remove* elements from this vector? We can do this by expanding on the indexing operations we learned earlier. Note that because we're working with the indexing operators `[]`, we'll be making references to the *indices* of vectors, not the values at those indices. For example, let's say we wanted to remove "Cambridge" from the vector. "Cambridge" is the third element, so you can run:
```{r}
example_vector[-3]
```
to remove that element.

**Exercise**: Did the line of code above *update* `example_vector`? e.g., does the saved value of `example_vector` still contain "Cambridge" in the 3rd position? Why or why not?

Here's an interesting question: what if you know that "Cambridge" is somewhere in the vector and you want to remove it, but you don't know where it is? You could solve this problem by running two commands: a `which` operation to find the index of "Cambridge", and a `[]` operation to remove that value:

```{r}
cambridge_index <- which(example_vector=="Cambridge") # note that I need to save this as a variable since I use it in the next line
example_vector[-cambridge_index]
```

This works, but there's actually an even faster way to do it: nest the `which` function *inside* your `[]` operator:

```{r}
example_vector[-which(example_vector=="Cambridge")]
```

In the same way as you have an "order of operations" for multiplication, R will interpret these nested functions in sequence: first it will run the `which` operation, then take the output (here, the number 3) and run the indexing operation on that number. Nesting R commands within each other is a crucial part of progamming-- it cuts down on the length of scripts, increases clarity, and often (once you get the hang of it) makes commands more readable to others. 

#### Combining vectors

Once again, let's consider our `example_vector`:

```{r, echo=F}
print(example_vector)
```

Now, let's say we have another string vector `second_vector`:
```{r}
second_vector <- c("Barcelona", "Madrid", "Cadiz")
print(second_vector)
```

How could we merge these two vectors into one? If you guessed that our combining function `c()` will be useful here, you were absolutly correct:

```{r}
merged_vector <- c(example_vector, second_vector)
print(merged_vector)
```
<br>

I know this is a lot of material to take in, but I assure you that with practice this will feel extremely natural. I'd recommend going through the first half of the problem set before continuing on to the "Logic and Loops" page.


## Side Note: Functions

I've used the term "function" a few times in this page, but haven't really defined it yet. A **function** is a piece of code that *performs an action* on other objects. If variables and arrays are the nouns of the coding world, functions are the verbs.

Functions have **arguments**, which are the input variables, and they produce **outputs**. Let's think back to that function `seq()`, which takes three arguments: a starting value, a stopping value, and the interval between values:
```{r}
seq(1, 10, 2)
```
<br>
The *output* of this function is a vector of values. But function outputs don't have to be vectors: they can be just about anything. The function `print()`, which we've seen several times before, doesn't strictly speaking "return" any output at all-- it just displays values on the screen. Other functions return single values, 2D arrays, or even other functions! 

Each argument in a function has a name, but you don't need to use them so long as you keep arguments in the same order that the function is expecting to see them. However, if you do choose to use argument names, you can list things in whatever order you like. These three lines of code are equivalent:
```{r}
seq(1, 10, 2)
#specify argument names, preserving original order
seq(from=1, to=10, by=2)
#specify argument names, scrambling order 
seq(to=10, by=2, from=1)
```
<br>
Notice that we use `=` rather than `<-` when passing named arguments to functions.

If you want to understand what a function does and what arguments it takes, type `?` followed by the function name into the console. This will bring the documentation for that function up in the `Help` window of RStudio. For instance, try typing `?seq` into the console and hitting enter. See what comes up? 

We won't get too deep into the nitty gritty of functions in this course, but I want you to have some idea of what they are and when you're using them. 




