---
title: "Troubleshooting"
output: html_document
---

Imagine yourself in a week. You've been plugging away at the problem set for a few hours, but your code just **is. not. working.** Everything you run leaves an error message or produces the wrong result and you have no idea what's wrong. You're so frustrated you want to throw your computer across the room. What can you do? 

The first three steps outlined below are things you can do on your own, without input from other people. Step 4 is where you start asking for help externally, and follows the format suggested in Module 1.

## Step 1: Learn to Love the Error Message 

Error messages are the best. Yes, they're red and scary, but error messages (when well-written) should point you to exactly the problem with your code. I've been programming for years, and I encounter error messages every day. It's nothing to be ashamed of.

How can we interpret error messages? Let's consider our broken code from a couple of pages ago:

```{r, error=T}
goldfish <- 8
print(godlfish)
```
<br> 
This is extremely useful information! The message tells us that our code is breaking because we're trying to perform an operation on an object that doesn't exist, and it's telling us the name of the problematic object! This makes it easy for us to see our mistake (a small typo), and correct it:
```{r}
print(goldfish)
```

Here's another example of a helpful error message:
```{r, error=T}
new_vector <- c("dog" "cat", "mouse")
```
<br> 
The `c()` function is expecting a series of elements separated by commas that it can group together into a vector. It was expecting a `,` after the word `"dog"`, but instead it got `"cat"`. This string was *unexpected*, hence the error message. Fix it by just adding a comma!


## Step 2: Chekc for Typos

Typos are by *far* the most common coding mistake. Probably 75% of the problems students brought to me last quarter involved some sort of small typo. Before you spend a lot of time online getting into the weeds of whatever function you're using, make sure you didn't just switch two letters around or forget a parenthesis. 

Speaking of forgetting parentheses, try typing this into your console and hitting Enter: `c("dog", "cat"`. Notice that instead of printing out the vector you've just specified, or giving you a `>` indicator like usual, you now have a `+` sign at the far left of your cursor. That means that you've opened some sort of bookended operator (like `""`, "()", or "{}"), but haven't closed it. The `+` indicates that the console is still waiting to get the end of its instruction. You can make it go away by actually closing the loop (in this case, by typing the missing `)`), or by pressing `Esc` to get back to the `>` cursor. 

RStudio trys to help avoid this issue by automatically generating the twin of most bookended operators when you type just one of them, but these types of mistake still happen. 

## Step 3: Check the Documentation

If you're having trouble with a function specifically, like `c()` or `seq()` or `print()`, the documentation for these functions is a great place to find some clarity. Every function that comes with R has a full documentation sheet specifying the name and purpose of that function, what arguments it takes, any details you should be aware of, and some example use cases. 

You can find this documentation online by searching for "{name of the function} R", but RStudio makes things even easier. Just type `?` followed by the name of the function, and the documentation should pop up in your `Help` panel. Try typing `?seq` into your console to get an example. 

The format of documentation can be a little nonintuitive at first, but you'll quickly get used to it. 

## Step 4: Ask the Internet: Googling

Programming broadly, and popular languages like R in particular, are fabulous because they have an incredibly vibrant online community that's constantly helping solve each other's problems. There are a lot of different forums for getting help with programming, but by far the most popular one is a site called (Stack Overflow)[www.stackoverflow.com]. The great thing about sites like these is that, especially for intro-level questions, you probably don't have to post a question yourself. You can just search for and read the answer to questions that have already been asked and answered. 

The idea of Googling problems we have is so obvious that it's become a clich&#232; ("Let me Google that for you"), but Googleing coding issues actually requires some skill and practice. The trick is in the lingo: a new programmer might search "putting a list in order from largest to smallest", whereas a more experienced one would search "sorting a vector in descending order". These syntactical differences can drastically affect the helpfulness of your search results. The only way to become a better Googler is to practice! 

### Tip 1: Google the Error Message

Yes, error messages are great, but sometimes they're a little cryptic. Googling the error message you got is a great way to quickly find forum posts addressing your specific bug. For example, our `print(godlfish)` command above generated the error message `Error in print(godlfish) : object 'godlfish' not found`. Try copy-pasting that into Google, with the letter "R" at the end to narrow things down to the R programming language. For me, the second-ranked output was an (extremely thorough explanation of this type of error message)[http://stackoverflow.com/questions/27886839/what-does-error-object-myvariable-not-found-mean]. 

Sometimes Google will get mixed up if you have unusual object names (like "goldfish") in your search query, so it's good to try and distill your error messages down to their essence: Googling "object not found error R" pulls up a whole suite of excellent Stack Overflow questions and R tutorials. 

### Tip 2: Google the Function Name

If `?{function name}` isn't helping you out, google the function (again, including "R" in your query somewhere). This will usually pull up loads of additional coding examples. 

### Tip 3: Google the Type of Issue

Are you trying to write a `for` loop, and just can't make heads or tail of it? Is my lecture not helping you gain clarity? The internet is full of additional tutorials that may be able to give you additional insights. 

## Step 5: Ask Your Peers

A lot of people taking this class already have some programming experience (and those who don't will quickly learn it). Posting to the discussion board is a wonderful way to get help from your classmates (and from Shrawan and me) in a way that also leaves a record of that problem in case someone else has that same issue later. 

There are very specific guidelines for writing a great help request. You can find them (here)[http://stackoverflow.com/help/how-to-ask] **Please follow them when posting!**

## Step 6: Email Me and/or Shrawan

I'm here if you need me, though I'd really prefer you post to the board so everyone can learn together. If you're emailing me with coding problems, please still **(follow the guidelines)[http://stackoverflow.com/help/how-to-ask] for asking a good question**! 

## Step 7: Ask the Internet: Writing your Own Stack Overflow Question

I suspect that nothing we do in this class will require you to ask a question that doesn't already have an answer online, but you never know! Posting to the discussion board  and sending me emails should leave you well-prepared to ask your own questions online. 


